---
title: "Practical 4 - Physiological Signals: Detecting Atrial Fibrillation from the ECG"
author: "Vivekkumar Dudhat"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
    theme: cosmo
    highlight: tango
    number_sections: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4,
  dpi        = 110,
  cache      = TRUE,
  cache.lazy = FALSE
)

GLOBAL_SEED <- 42
set.seed(GLOBAL_SEED)
```

> **Submission.** Knit your Rmd file into HTML and submit it via
> Brightspace. Include a link to your git repository (GitHub or
> git.cs.dal.ca) as a comment on your submission.

------------------------------------------------------------------------

## 0 Setup

This lab uses a few different/new libraries to handle signal processing
and to cover all 3 of our analysis paradigms for signal data: a
**time-domain** model, a **frequency-domain** model, and a
**state-space** model.

```         
install.packages(c(
  "signal",      # digital filters (Butterworth band-pass / notch)
  "forecast",    # ARIMA time-series models (auto.arima)
  "depmixS4"     # Gaussian (continuous) hidden Markov models (state-space)
))
```

If the block below runs without error you are ready to go. Instead of
attaching like normal (using `library()`) we are just checking if
`signal` and `forecast` are both installed (using `requiredNamespace`).
This is these libraries contain functions (like `filter()`) that collide
with each other and `dplyr`. Instead we will call both explicitly in the
code with the `signal::` / `forecast::` prefix where needed.

```{r packages}
library(tidyverse)
library(here)
library(fs)
library(depmixS4)
library(yardstick)
library(scales)
requireNamespace('signal')
requireNamespace('forecast')
```

------------------------------------------------------------------------

## 1 Introduction

The electrocardiogram (ECG) records the small voltages produced as the
heart's chambers depolarise to contract (systole) and repolarise to
relax (diastole).

A single normal beat has a very distinctive/standard shape - the **P
wave** (atrial contracting), the **QRS complex** (ventricular
contracting), and the **T wave** (ventricular relaxing). The time
between the main **R** spikes in successive QRS peaks is the **R-R
interval**.

When a heart doesn't beat in a nice regular rhythm we call this a
cardiac arrhythmia. **Atrial fibrillation (AF)** is the most common
sustained arrhythmia and a major cause of stroke and heart failure. As
AF is often asymptomatic, a lot of work has gone into identifying it
accurately from ECGs. Typically, AF is recognised in an ECG by the
disappearance of the **P wave** (as the atria are contracting
chaotically rather than in an organised wave) and irregular spacing and
timing of the **QRS** complex (as the ventricles contract at irregular
intervals).

In this lab we will build a different AF detector for each signal data
analysis paradigm and then compare them against one another:

- a **time-domain** model that asks how predictable the next R-R
  interval is (using **ARIMA** via `forecast::auto.arima`, the
  autoregressive / moving-average / integrated family from the lecture),
- a **frequency-domain** model that asks how the signal's energy is
  spread across frequencies (using the Fourier transform), and
- a **state-space** model that treats the rhythm as a sequence of hidden
  states (a hidden Markov model).

## 2 The Data

We use a small prepared subset of the **MIT-BIH Atrial Fibrillation
Database** (the same public PhysioNet database Babaeizadeh et al.
validated on): two-channel Holter ECGs sampled at 250 Hz with expert
beat and rhythm annotations. Reading the raw PhysioNet files needs some
slightly awkward platform-specific libraries, so I've simplified things
a little (including extract R-R intervals) and converted to a more
usable format for you. We've also simplified away the "temporal
dependence" complexity of how you cut-up time-series data to avoid
misleading yourself.

As in the earlier labs, we download a prepared archive once and cache
it.

```{r data-paths}
data_dir  <- here("data")
cache_dir <- fs::path(data_dir, "cache")
fs::dir_create(cache_dir)

data_url <- "https://maguire-lab.github.io/health_data_science_research_2026/static_files/practicals/ecg_afib.zip"
zip_path <- fs::path(cache_dir, "ecg_afib.zip")
ecg_root <- fs::path(data_dir, "ecg_afib")
```

```{r download-data, cache = TRUE}
if (!fs::dir_exists(ecg_root)) {
  if (!fs::file_exists(zip_path)) {
    download.file(data_url, zip_path, mode = "wb", quiet = TRUE)
  }
  unzip(zip_path, exdir = data_dir)
}

fs::dir_ls(ecg_root)
```

The archive contains 2 tables:

- `waveforms.csv` - a handful of 10-second raw ECG snippets (2500
  samples each at 250 Hz), labelled `AF` or `Normal` where `record`
  indicates an individual patient.
- `rr_intervals.csv` - sequences of consecutive R-R intervals split into
  fixed-length **windows** (128 beats each, roughly two minutes) and
  labelled by rhythm.

```{r load-data}
waveforms <- read_csv(fs::path(ecg_root, "waveforms.csv")) |>
  mutate(rhythm = factor(rhythm, levels = c("AF", "Normal")))

rr <- read_csv(fs::path(ecg_root, "rr_intervals.csv")) |>
  mutate(rhythm = factor(rhythm, levels = c("AF", "Normal")))

glimpse(waveforms)
glimpse(rr)
```

**Q1 (2 marks):** Write a short data dictionary for `waveforms` and `rr`
with your best guess explanation of the values and format of these
datasets.

**Answer Q1:**

| Column | Type | Description |
|---------------------|------------------|---------------------------------|
| `record` | string | Patient identifier from the MIT-BIH AF Database. There are 6 unique patients: `04015`, `04746`, `05121`, `06426`, `07879`, `08378`. |
| `segment` | string | Unique identifier for a 10-second ECG snippet, formed as `<record>_<rhythm>` (e.g. `04015_Normal`). Each segment is one continuous clip from a single patient and rhythm. |
| `rhythm` | factor | Expert-annotated rhythm label for the segment `"AF"` (atrial fibrillation) or `"Normal"` (sinus rhythm). Balanced: 15,000 rows per class. |
| `time` | numeric | Time in seconds from the start of the segment (0.000 to 9.996 s, step = 0.004 s), reflecting the 250 Hz sampling rate. |
| `ecg` | numeric | ECG voltage in millivolts (mV), ranging approximately 3.37 to +3.41 mV. Represents the bioelectrical signal recorded at the electrode capturing the heart's depolarisation and repolarisation cycles. |

| Column | Type | Description |
|---------------------|------------------|---------------------------------|
| `record` | string | Same patient identifier as above (6 patients). |
| `window` | integer | Index (0-based) of a non-overlapping 128-beat window within a record. Each window covers roughly two minutes of heartbeats and carries a single rhythm label. |
| `rhythm` | factor | Expert-annotated rhythm label for the window `"AF"` or `"Normal"`. Imbalanced: approximately 34,000 AF beat rows versus 21,500 Normal beat rows. |
| `beat` | integer | Beat number (0 based) within the window, ranging from 0 to 127. |
| `rr` | numeric | R-R interval in seconds — the time elapsed between two successive R-peaks. Values range from approximately 0.20 s to 1.99 s (roughly 30–300 bpm). Intervals are shorter and highly irregular in AF; longer and more regular in Normal sinus rhythm. |

## 3 Train / Test Split

Before any cleaning or exploration we split the data, so that nothing we
decide later can be influenced by the test set.

We split on whole **records** and keep the train and test data in
completely separate objects (`rr_train` / `rr_test` and
`waveforms_train` / `waveforms_test`).

```{r split}
set.seed(GLOBAL_SEED)

all_records  <- rr |> distinct(record) |> pull(record)
test_records <- sample(all_records, size = round(0.3 * length(all_records)))

rr_train <- rr |> filter(!record %in% test_records)
rr_test  <- rr |> filter(record %in% test_records)

waveforms_train <- waveforms |> filter(!record %in% test_records)
waveforms_test  <- waveforms |> filter(record %in% test_records)

bind_rows(
  rr_train |> distinct(record, window, rhythm) |> mutate(split = "train"),
  rr_test  |> distinct(record, window, rhythm) |> mutate(split = "test")
) |>
  count(split, rhythm)
```

**Q2 (2 marks):** Why do we split on the **record** instead of the
**segment**? What could the models end up learning to recognise instead
of AF if we split by **segment**?

**Answer Q2:** We split by **record** (patient) rather than segment
because each patient contributes multiple segments and windows. If we
split by segment, the same patient's data would appear in both the
training and test sets. Since every individual has a unique ECG
signature their own baseline voltage level, QRS morphology, T-wave
shape, and heart rate range. model could learn to identify the *patient*
rather than the rhythm. It would then perform well on the test set not
because it has learned what AF looks like, but because it has
effectively memorised that patient's signal. This is a form of **data
leakage**: the test set is no longer independent, and the reported
performance is overly optimistic and would not generalise to new
patients.

Specifically, if split by segment, the models could learn to recognise
individual patient identity (e.g., a distinctive QRS amplitude or
baseline offset), or they could exploit the fact that AF and Normal
segments from the same patient are recorded under similar conditions
(same electrode placement, same recording session), meaning differences
between classes are confounded with within-patient variation rather than
true rhythm differences.

## 4 Cleaning - Frequency Filtering

Raw ECG is rarely analysis-ready. Two classic problems are **baseline
wander** (slow drift from breathing and electrode movement, below the
heartbeat frequency) and **mains interference**(a sharp hum at the
power-line frequency - 60 Hz as these are US recordings - [table of
national
frequencies](https://www.wpowerproducts.com/power-generation-resources/voltages-and-frequencies-by-country/)).

The lecture covered two standard fixes: a **band-pass** filter keeps
only frequencies likely to be connected to heart bioelectical signals,
and a **notch (band-stop)** filter removes narrow bands of interference
like the **mains frequency**. The the actual way these filters are
implemented varies a lot (with different nuances on how they
flatten/amplify things). In this lab we implement the two filters using
something called a [Butterworth
filter](https://en.wikipedia.org/wiki/Butterworth_filter) from the
`signal` and apply them using `filtfilt()` (which filters forwards and
backwards so features are not shifted in time).

```{r filtering, fig.width = 8, fig.height = 4.5}
fs <- 250

# keep things in conceivable "normal heart rate" range (0.5 Hz to 40 Hz) to smooth out high and low frequency noise 
bandpass <- signal::butter(2, c(0.5, 40) / (fs / 2), type = "pass")

# cut out 59-61Hz signal from the electrical grid
notch    <- signal::butter(2, c(59, 61) / (fs / 2), type = "stop")

example_wave <- waveforms_train |>
  filter(rhythm == "Normal") |>
  filter(segment == first(segment))

cleaned <- example_wave |>
  mutate(
    bp_filtered = signal::filtfilt(bandpass, ecg),
    filtered    = signal::filtfilt(notch, bp_filtered)
  )

cleaned |>
  filter(time <= 4) |>
  dplyr::select(time, Raw = ecg, Filtered = filtered) |>   # dplyr::select - MASS also exports select()
  pivot_longer(c(Raw, Filtered), names_to = "stage", values_to = "ecg") |>
  mutate(stage = factor(stage, levels = c("Raw", "Filtered"))) |>
  ggplot(aes(time, ecg)) +
  geom_line(linewidth = 0.3) +
  facet_wrap(~stage, ncol = 1) +
  labs(x = "Time (s)", y = "ECG (mV)",
       title = "Before and after band-pass + notch filtering")
```

As we can see, this cleans things up a little (although isn't perfect)
so now let's apply the **same** two filters to every waveform in both
the training and test sets.

We keep the original trace as `ecg_raw` and overwrite `ecg` with the
cleaned signal.

```{r apply-filter, cache = TRUE}
clean_waveforms <- function(df) {
  df |>
    group_by(record, segment) |>
    arrange(time, .by_group = TRUE) |>
    mutate(
      ecg_raw = ecg,
      ecg     = signal::filtfilt(notch, signal::filtfilt(bandpass, ecg))
    ) |>
    ungroup()
}

waveforms_train <- clean_waveforms(waveforms_train)
waveforms_test  <- clean_waveforms(waveforms_test)

glimpse(waveforms_train)
```

**Q3 (1 marks):** As part of these filters we are dividing the sampling
rate in half. What does fs/2 represent in terms of the signal we've
captured?

**Answer Q3:** `fs/2` is the **Nyquist frequency** — the highest
frequency that can be faithfully represented in a signal sampled at rate
`fs`. At 250 Hz, `fs/2 = 125 Hz`. By the Nyquist theorem, to capture a
frequency without aliasing (where high-frequency content folds back and
appears as a false lower frequency), you must sample at least twice as
fast as that frequency. Dividing filter cutoffs by `fs/2` normalises
them to the range [0, 1], where 1 corresponds to the absolute maximum
frequency the recording can contain. So a band-pass cutoff of
`0.5 / (fs/2)` means "keep frequencies above 0.5 Hz" and `40 / (fs/2)`
means "keep frequencies below 40 Hz" both well within the 125 Hz limit
of this 250 Hz recording.

## 5 Exploratory Data Analysis

Remember, we only explore the **training set**!

### 5.1 Class balance and a first look

```{r eda-balance, fig.height = 3}
rr_train |>
  distinct(record, window, rhythm) |>
  count(rhythm) |>
  ggplot(aes(rhythm, n, fill = rhythm)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("AF" = "firebrick", "Normal" = "grey60")) +
  labs(x = NULL, y = "Number of windows (train)", title = "Window class balance")
```

A normal and an AF waveform snippet side by side - note the slightly
more regular spacing in the Normal vs AF signal; the p-wave feature is
less obviously different in this case:

```{r eda-waveforms, fig.width = 8, fig.height = 4}
waveforms_train |>
  filter(time <= 4) |>
  group_by(rhythm) |>
  filter(segment == first(segment)) |>
  ungroup() |>
  ggplot(aes(time, ecg)) +
  geom_line(linewidth = 0.3) +
  facet_wrap(~rhythm, ncol = 1) +
  labs(x = "Time (s)", y = "ECG (mV)", title = "Normal vs AF (training examples)")
```

### 5.2 Heart-rate-variability features

A few standard summaries of each window's R-R series:

```{r eda-hrv}
window_features <- rr_train |>
  group_by(record, window, rhythm) |>
  summarise(
    mean_rr = mean(rr),                       # average interval (1 / mean_rr = mean rate)
    sdnn    = sd(rr),                          # overall variability
    .groups = "drop"
  )

window_features |>
  pivot_longer(c(mean_rr, sdnn),
               names_to = "feature", values_to = "value") |>
  ggplot(aes(rhythm, value, fill = rhythm)) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(~feature, scales = "free_y") +
  scale_fill_manual(values = c("AF" = "firebrick", "Normal" = "grey60")) +
  labs(x = NULL, y = NULL, title = "HRV features by rhythm (train)")
```

**Q4 (2 marks):** Explain what this plot tells you about the R-R
intervals in Normal vs AF samples.

**Answer Q4:** The plot shows two boxplots — `mean_rr` (average R-R
interval per window) and `sdnn` (standard deviation of R-R intervals per
window) — split by rhythm.

For **`mean_rr`**, the AF and Normal distributions overlap considerably,
though AF windows tend to have slightly shorter mean intervals on
average, reflecting the faster and more chaotic ventricular rate typical
of AF. However, the overlap means `mean_rr` alone is not a reliable
separator of the two classes.

For **`sdnn`**, there is a much clearer separation: AF windows have
substantially higher standard deviations than Normal windows. This
directly captures the hallmark of AF — the highly irregular beat-to-beat
timing. Normal sinus rhythm intervals cluster tightly around a
consistent mean (low SDNN), while AF intervals jump erratically from
beat to beat (high SDNN). The `sdnn` feature is therefore far more
discriminative than `mean_rr` for distinguishing AF from Normal, and
this irregularity is exactly what all three models in this lab aim to
exploit.

## 6 Time Domain Model (ARIMA)

To detect AF with a time-domain approach we are going to work out **how
predictable is the next interval given the preceding one(s) across each
window?**. A regular sinus rhythm is highly predictable, so our model
should be able to forecast the next intervals well; whereas an AF window
is close to noise, so the forecast error is large. This means we can use
our model's **forecast error as the AF score**.

We'll use an ARIMA model to do this. ARIMA predicts the next value from
a linear combination of recent values (the AR - autoregressive part) and
recent forecast errors (the MA - moving average part), after
differencing the series to make it stationary (the I part).\
Specifically, we are going to use `forecast::auto.arima()` to
automatically select (hyperparameter optimisation) how many previous
values to use for each part of this model (also known as the "order" of
the model component).

For each window we treat the R-R series as a tachogram (interval value
indexed by beat number), fit ARIMA on the first 80% of beats, forecast
the rest, and record the mean absolute error.

```{r arima-helper}
arima_error <- function(window_df) {
  y   <- window_df$rr
  h   <- max(5, round(0.2 * length(y)))   # forecast horizon = last 20% of beats
  cut <- length(y) - h

  fit      <- forecast::auto.arima(y[seq_len(cut)])
  forecast <- forecast::forecast(fit, h = h)
  mean(abs(as.numeric(forecast$mean) - y[(cut + 1):length(y)]))   # forecast MAE
}
```

Here is the idea on one window of each rhythm - the red forecast tracks
the regular rhythm closely but cannot keep up with AF:

```{r arima-demo, fig.width = 8, fig.height = 3.4}
demo <- rr_train |>
  group_by(rhythm) |>
  filter(window == last(window)) |>
  ungroup()

demo_forecast <- demo |>
  group_by(rhythm) |>
  group_modify(~ {
    y <- .x$rr
    h <- max(5, round(0.2 * length(y)))
    cut <- length(y) - h
    fit <- forecast::auto.arima(y[seq_len(cut)])
    fc  <- forecast::forecast(fit, h = h)
    tibble(beat     = seq_along(y),
           rr       = y,
           forecast = c(rep(NA_real_, cut), as.numeric(fc$mean)))
  }) |>
  ungroup()

# forecast error (mean absolute error over the forecast horizon) for each rhythm
demo_error <- demo_forecast |>
  filter(!is.na(forecast)) |>
  group_by(rhythm) |>
  summarise(mae = mean(abs(forecast - rr)), .groups = "drop") |>
  mutate(label = paste0("MAE = ", round(mae, 3), " s"))

ggplot(demo_forecast, aes(beat)) +
  geom_line(aes(y = rr), linewidth = 0.3) +
  geom_point(aes(y = rr), size = 0.5) +
  geom_line(aes(y = forecast), colour = "firebrick", linewidth = 0.6, na.rm = TRUE) +
  geom_text(data = demo_error, aes(x = Inf, y = Inf, label = label),
            hjust = 1.1, vjust = 1.5, size = 3.5, inherit.aes = FALSE) +
  facet_wrap(~rhythm, scales = "free_x") +
  labs(x = "Beat number", y = "R-R interval (s)",
       title = "ARIMA forecast (Red) vs Actual")
```

We build a per-window key table and a matching list of windows for the
**train** and **test** sets separately. Each model below scores both
sets - we look at the training scores to sanity-check that the score
separates the classes, and keep the test scores untouched/unviewed for
the final evaluation at the end. Now score every window. This fits one
ARIMA model per window; it is cached so it only re-runs if the data
changes.

```{r window-setup, cache = TRUE}
train_grp  <- rr_train |> group_by(record, window, rhythm)
train_keys <- group_keys(train_grp)
train_list <- group_split(train_grp)

test_grp  <- rr_test |> group_by(record, window, rhythm)
test_keys <- group_keys(test_grp)
test_list <- group_split(test_grp)

train_keys$arima_err <- map_dbl(train_list, arima_error)
test_keys$arima_err  <- map_dbl(test_list, arima_error)

```

Now we can compare the forecast error in Normal vs AF across our
training set:

```{r arima-scores-show}
train_keys |>
  group_by(rhythm) |>
  summarise(mean_forecast_error = mean(arima_err), .groups = "drop")
```

The mean error is higher for AF, but how well does the forecast error
*alone* separate AF from Normal? We are treating `arima_err` directly as
an AF score (larger = more AF-like). To make a prediction of AF vs
Normal we to pick a threshold score i.e., if `arima_err` is greater than
threshold we predict AF and below Normal.

To do this we look at the **ROC curve**, which traces sensitivity
against 1 - specificity as we sweep the detection threshold across every
value of `arima_err`. We then pick the threshold value that maximises
the Area Under the Curve (or **AUC**).

```{r arima-roc, fig.width = 5, fig.height = 5}
train_keys |>
  roc_curve(truth = rhythm, arima_err, event_level = "first") |>
  autoplot() +
  labs(title = "ROC: ARIMA forecast error as an AF detector (train)")

train_keys |>
  roc_auc(truth = rhythm, arima_err, event_level = "first")
```

**Q5 (1 marks):** What other type of curve could we use to select the
threshold? (Hint: think back to earlier labs)

**Answer Q5:** We could use a **precision-recall (PR) curve**, which
plots precision (positive predictive value) against recall (sensitivity)
across all possible thresholds. Unlike the ROC curve, the PR curve is
more informative under class imbalance. It directly shows the trade-off
between how many predicted AF cases are correct (precision) and how many
true AF cases are caught (recall). The threshold that maximises the **F1
score** (the harmonic mean of precision and recall) is a common
operating point choice from the PR curve.

## 7 Frequency Domain

The frequency-domain formulation we are going to use is **does AF and
Normal signals have distinguishable amounts of different frequencies in
their signals?**

We can infer the frequency distribution of a signal using the Fourier
transform (base R's `fft()`). A regular rhythm varies slowly (its R-R
series wanders gently with breathing), so we expect its energy sits at
*low* frequencies in a single concentrated peak. AF jumps from beat to
beat, so its energy spreads out and up into *higher* frequencies. Rather
than collapse this into one hand-picked score, we summarise each
window's spectrum with **a handful of frequency-space features** and
then let a **logistic regression** learn how to weight them.

For each window we compute four interpretable frequency statistics:

- `hf_frac` - fraction of power in the **upper half** of the band (AF
  pushes power up here),
- `spectral_centroid` - the power-weighted **mean frequency bin** (the
  spectrum's "centre of mass"; higher for AF),
- `spectral_entropy` - how **spread out / noisy** the spectrum is,
  normalised to `[0, 1]` (a flat, broadband AF spectrum scores near 1; a
  single sharp peak scores near 0),
- `peak_frac` - the share of power in the **single biggest bin** (a
  regular rhythm concentrates power in one peak, so this is high for
  Normal).

```{r freq-features}
freq_features <- function(rr) {
  x     <- rr - mean(rr)                       # remove the DC offset
  n     <- length(x)
  power <- (Mod(fft(x))^2)[2:floor(n / 2)]     # one-sided power, dropping the DC term
  if (sum(power) == 0) {
    return(tibble(hf_frac = 0, spectral_centroid = 0,
                  spectral_entropy = 0, peak_frac = 0))
  }
  k    <- length(power)
  bins <- seq_len(k)
  p    <- power / sum(power)                    # spectrum as a probability distribution

  tibble(
    hf_frac           = sum(power[round(k / 2):k]) / sum(power),
    spectral_centroid = sum(bins * p),
    spectral_entropy  = -sum(p * log(p + 1e-12)) / log(k),
    peak_frac         = max(p)
  )
}

# attach the features to each window's key table. Assigning by column name
# overwrites any existing copies, so re-running this block never duplicates or
# mangles column names (unlike bind_cols).
train_feats <- map_dfr(train_list, ~ freq_features(.x$rr))
test_feats  <- map_dfr(test_list,  ~ freq_features(.x$rr))

train_keys[names(train_feats)] <- train_feats
test_keys[names(test_feats)]   <- test_feats

train_keys |>
  group_by(rhythm) |>
  summarise(across(c(hf_frac, spectral_centroid, spectral_entropy, peak_frac), mean),
            .groups = "drop")
```

Now we fit a simple logistic regression that predicts the probability a
window is **AF** from those four features, using the **training**
windows only. The fitted probability is the frequency model's AF score
(`freq_prob`).

```{r freq-logreg}
freq_model <- glm(
  (rhythm == "AF") ~ hf_frac + spectral_centroid + spectral_entropy + peak_frac,
  family = binomial,
  data   = train_keys
)

summary(freq_model)

# predicted probability of AF for every window
train_keys$freq_prob <- predict(freq_model, type = "response")
test_keys$freq_prob  <- predict(freq_model, newdata = test_keys, type = "response")
```

**Q6 (2 marks)** Based on the output from summary above which
frequency-based features are the best predictors of AF in this dataset?
Does this match our hypotheses (above) for what each of these features
will capture?

**Answer Q6:** In the logistic regression summary, the best predictors
are identified by the smallest p-values and the largest absolute
z-values. Based on both the model output and the hypotheses laid out
above:

- **`spectral_entropy`** is the strongest predictor. It has a large
  positive coefficient and a highly significant p-value, reflecting that
  AF windows have a broadly spread, near-uniform spectrum (entropy near
  1) while Normal windows concentrate power in a single peak (entropy
  near 0). This matches the hypothesis exactly.
- **`peak_frac`** is also a strong predictor with a large negative
  coefficient. Normal rhythms concentrate most power in one dominant
  frequency bin (high `peak_frac`), whereas AF spreads it across many
  bins (low `peak_frac`). This is effectively the mirror image of
  `spectral_entropy` and also matches the hypothesis.
- **`hf_frac`** and **`spectral_centroid`** tend to be less significant
  or only marginally significant. This is partly because AF pushes some
  energy to higher frequencies, but the effect is weaker and partially
  captured by `spectral_entropy` already (the features are correlated).
  The hypothesis that AF shifts energy upward is directionally correct
  but these features add less unique information once entropy and peak
  fraction are in the model.

Overall the results match the hypotheses well: the features that most
directly capture rhythm irregularity (entropy, peak concentration)
outperform those that capture frequency location (centroid,
high-frequency fraction).

Similarly to the time-based model above, we have to pick a threshold on
the logistic regression output to decide when to make a prediction of
AF. We can sweep possible thresholds from 0 to 1 to again trace out the
**ROC curve**. Each point is one cut-off for calling a window AF; the
dashed diagonal is chance, and the more the curve bows towards the
top-left the better the model separates AF from Normal:

```{r freq-roc, fig.width = 5, fig.height = 5}
train_keys |>
  roc_curve(truth = rhythm, freq_prob, event_level = "first") |>
  autoplot() +
  labs(title = "ROC: logistic-regression frequency detector (train)")

train_keys |>
  roc_auc(truth = rhythm, freq_prob, event_level = "first")

```

**Q7 (1 mark)**: What does it mean for the ROC curve to be below the
dotted line?

**Answer Q7:** The dotted diagonal line on a ROC curve represents the
performance of a **random classifier** (AUC = 0.5) — one that assigns
scores with no relationship to the true class. A curve that falls
*below* this line means the model is performing **worse than random
chance**: it is systematically assigning higher scores to Normal windows
and lower scores to AF windows, i.e., the scoring direction is inverted.
This does not mean the model has learned nothing. it has actually
learned a real signal, but got the direction backwards. Flipping all
predictions (predicting AF when the score is low and Normal when it is
high) would produce a curve that mirrors above the diagonal and would
perform better than random.

## 8 State Space (Hidden Markov Model)

The state-space model formulation is: **what sequence of hidden states
best explains the observed intervals?** where the rhythm moves between
unobserved states that each emit specific intervals with different
probabilities.

Rather than discretising each interval into a symbol, we model the
**interval values directly** as continuous series of emitted values. We
fit a two-state HMM on all the training windows using `depmixS4`. With
two hidden states the model tends to discover one "regular" state
(intervals tightly clustered around a mean) and one "irregular" state (a
broad, high-variance distribution). Crucially we never show it the
AF/Normal labels - the states in this case are learned purely from the
distribution of R-R intervals.

```{r hmm-train, cache = TRUE}
# concatenate the training windows into one series, recording each window's length
# so depmixS4 treats the windows as separate sequences (the ntimes argument)
train_seq    <- bind_rows(train_list)
ntimes_train <- map_int(train_list, nrow)

hmm_spec <- depmix(rr ~ 1, data = train_seq, nstates = 2,
                   family = gaussian(), ntimes = ntimes_train)
hmm_fit  <- fit(hmm_spec, verbose = FALSE)

summary(hmm_fit)
```

The two states differ in their emission distributions - one has a much
larger standard deviation (the AF-like, irregular state). We use the
**training labels** only to decide *which* hidden state that is: the
state that accounts for most of the beats occurring in AF windows.

```{r hmm-af-state}
train_states <- posterior(hmm_fit, type = "viterbi")$state   # decoded state per beat

af_state <- tibble(rhythm = train_seq$rhythm, state = train_states) |>
  count(state, rhythm) |>
  group_by(state) |>
  summarise(af_frac = sum(n[rhythm == "AF"]) / sum(n), .groups = "drop") |>
  slice_max(af_frac, n = 1, with_ties = FALSE) |>
  pull(state)
```

Now to label a whole **window** we infer the underlying hidden states
across the window and look at how much of it sits in the AF state. The
*fraction* of a window's beats in the AF state is a natural continuous
AF score; a simple majority vote would threshold it at 0.5, but below we
instead fit the best threshold from an ROC curve (below).

For new (test) data, `depmixS4` predicts by building a model with the
same structure, copying the fitted parameters into it, and decoding - it
does not re-estimate anything:

```{r hmm-score}
# Viterbi-decode a list of windows under an already-fitted model
decode_states <- function(window_list, fitted) {
  seq_df <- bind_rows(window_list)
  spec   <- depmix(rr ~ 1, data = seq_df, nstates = 2,
                   family = gaussian(), ntimes = map_int(window_list, nrow))
  spec   <- setpars(spec, getpars(fitted))   # reuse the trained parameters
  posterior(spec, type = "viterbi")$state
}

# fraction of each window's beats that fall in the AF state
af_fraction <- function(states, window_list) {
  idx <- rep(seq_along(window_list), map_int(window_list, nrow))
  as.numeric(tapply(states == af_state, idx, mean))
}

train_keys$hmm_score <- af_fraction(train_states, train_list)
test_keys$hmm_score  <- af_fraction(decode_states(test_list, hmm_fit), test_list)
```

Now let's fit the threshold we are going to use to determine how much of
a window should be in the AF state to classify the window as AF.

```{r hmm-roc, fig.width = 5, fig.height = 5}
roc_hmm <- train_keys |>
  roc_curve(truth = rhythm, hmm_score, event_level = "first")

hmm_threshold <- roc_hmm |>
  mutate(youden = sensitivity + specificity - 1) |>
  slice_max(youden, n = 1, with_ties = FALSE)

hmm_threshold

autoplot(roc_hmm) +
  geom_point(data = hmm_threshold,
             aes(x = 1 - specificity, y = sensitivity),
             colour = "firebrick", size = 3) +
  labs(title = "ROC: HMM AF-state fraction as a detector (train)",
       subtitle = paste0("Chosen threshold (max Youden's J) = ",
                         round(hmm_threshold$.threshold, 3)))
```

We then label every window by comparing its AF-state fraction to that
fitted threshold:

```{r hmm-predict}
label_hmm <- function(df) {
  df |> mutate(hmm_pred = factor(if_else(hmm_score >= hmm_threshold$.threshold,
                                         "AF", "Normal"),
                                 levels = c("AF", "Normal")))
}
train_keys <- label_hmm(train_keys)
test_keys  <- label_hmm(test_keys)
```

How well does that rule separate the rhythms on the training windows?

```{r hmm-confusion}
train_keys |>
  conf_mat(truth = rhythm, estimate = hmm_pred)
```

**Q8 (2 marks):** In this HMM, what is *observed* and what is *hidden*?

**Answer Q8:** In this HMM, the **observed** variable is the sequence of
**R-R interval values** (`rr`) — the actual measured times in seconds
between consecutive heartbeats. These are the numbers the model directly
receives as input for each window. The **hidden** variable is the
**underlying rhythm state** at each beat one of two discrete latent
states (state 1 or state 2) that the model infers but never directly
observes. The model learns that one state emits R-R intervals drawn from
a tight Gaussian distribution (low mean, low variance corresponding to a
regular, Normal sinus rhythm) and the other state emits intervals from a
broader Gaussian (higher variance corresponding to the irregular
beat-to-beat timing of AF). Crucially, the model is never given the
AF/Normal labels during training; it discovers these two states purely
from the shape of the interval distributions. The rhythm labels are only
used *after* fitting, to determine which of the two learned states
corresponds to AF.

## 9 Comparing the Three Detectors

Each model gives every window a score where larger means more AF-like.
To compare them fairly we rescale each score to `[0, 1]` (so they share
an axis) and evaluate on the **held-out test records**. The **ROC
curve** and its **AUC** summarise how well each score separates AF from
normal across all possible cut-offs - the same yardstick both papers
report.

```{r compare, fig.width = 6, fig.height = 4.5}
test_scored <- test_keys |>
  mutate(across(c(arima_err, freq_prob, hmm_score), rescale)) |>   # higher = more AF
  pivot_longer(c(arima_err, freq_prob, hmm_score),
               names_to = "model", values_to = "score") |>
  mutate(
    model = recode(model,
                   arima_err = "Time (ARIMA)",
                   freq_prob = "Frequency (logistic regression)",
                   hmm_score = "State-space (HMM)"),
    truth = factor(rhythm, levels = c("AF", "Normal"))
  )

# ROC curves, all three models on one set of axes
test_scored |>
  group_by(model) |>
  roc_curve(truth, score, event_level = "first") |>
  autoplot() +
  labs(title = "ROC: three AF detectors on the test set")
```

```{r compare-auc}
test_scored |>
  group_by(model) |>
  roc_auc(truth, score, event_level = "first") |>
  dplyr::select(model, .estimate) |>   # dplyr::select - MASS (via depmixS4) also exports select()
  arrange(desc(.estimate))
```

**Q9 (2 marks):** How do these models compare to one another?

**Answer Q9:** Comparing the three models on the held-out test set using
AUC (from the `compare-auc` chunk, arranged in descending order):

- The **State-space (HMM)** model typically achieves the highest AUC.
  This makes sense because it explicitly models the *sequential
  structure* of R-R intervals. it asks which hidden state each beat
  belongs to and uses the full temporal pattern of state transitions
  across the window rather than a single summary statistic.
- The **Frequency-domain (logistic regression)** model performs strongly
  as well. By capturing four complementary spectral features (entropy,
  peak concentration, centroid, high-frequency fraction), it summarises
  the broadband irregularity of AF effectively. Its performance is close
  to the HMM because spectral entropy and peak fraction are highly
  informative on their own.
- The **Time-domain (ARIMA)** model generally achieves a lower AUC than
  the other two. While the forecast error does separate AF from Normal
  (AF windows are harder to predict), ARIMA reduces each window to a
  single scalar (mean absolute forecast error) and may not fully capture
  the nature of AF irregularity compared to the feature-rich frequency
  model or the sequence-aware HMM.

All three models should comfortably outperform the random baseline (AUC
= 0.5), demonstrating that each of the three signal analysis paradigms
time-domain, frequency-domain, and state-space captures meaningful
information about AF rhythm.

**Q10 (3 marks):** Suggest a way we could improve each of these models
performance and suggest at least 3 additional analyses that it would be
a good idea to perform before trying to use these in the real-world.

**Answer Q10:**

**Model improvements:**

- **Time-domain (ARIMA):** Rather than collapsing a whole window to a
  single forecast MAE, we could extract richer time-domain HRV features
  per window such as RMSSD (root mean square of successive differences),
  pNN50 (proportion of successive intervals differing by more than 50
  ms), and the coefficient of variation and feed them into a logistic
  regression or random forest. These are established clinical HRV
  metrics that more completely characterise beat-to-beat irregularity
  than ARIMA forecast error alone.

- **Frequency-domain (logistic regression):** The current features are
  all derived from the R-R interval tachogram. A direct improvement
  would be to compute frequency features from the **raw ECG waveform**
  (using `waveforms`) rather than the R-R series, which would capture
  the loss of the P wave a hallmark of AF alongside spectral
  irregularity. Additionally, replacing logistic regression with a more
  powerful classifier (e.g., random forest or gradient boosted trees)
  would allow non-linear interactions between features to be exploited.

- **State-space (HMM):** The current model uses only two states and a
  single observed variable (`rr`). Performance could be improved by
  increasing the number of states to capture transitional rhythms, or by
  building a **multivariate HMM** that jointly models R-R intervals
  alongside additional features (e.g., QRS duration or amplitude derived
  from the waveform). A deep learning alternative such as an LSTM, which
  models long-range sequential dependencies, would likely outperform a
  Gaussian HMM on longer recordings.

**Additional real-world analyses:**

1.  **External validation on an independent dataset.** All three models
    were trained and tested on a single database (MIT-BIH AFIB). Before
    clinical use, they should be evaluated on data from a different
    institution, country, or device type to assess whether performance
    generalises beyond the specific recording conditions, patient
    population, and annotation protocol of this dataset.

2.  **Subgroup analysis.** AF presentation and ECG morphology can differ
    by age, sex, and comorbidities (e.g., heart failure, diabetes).
    Performance should be broken down by patient subgroup to check that
    no group is systematically disadvantaged a model with strong overall
    AUC can still perform poorly on elderly patients or those with
    co-existing arrhythmias.

3.  **Clinical operating point and calibration analysis.** AUC
    summarises performance across all thresholds but clinicians need to
    operate at one specific threshold. A full sensitivity/specificity
    analysis at clinically meaningful operating points (e.g., targeting
    ≥ 90% sensitivity for a screening tool) is needed, along with
    calibration curves to verify that predicted probabilities actually
    reflect true AF likelihood a model can have high AUC but poorly
    calibrated probabilities, which matters if the score is used to
    triage urgency of referral.

------------------------------------------------------------------------

## Reproducibility

```{r session-info, cache = FALSE}
sessionInfo()
```
